On Devon Allen's Disqualification at the 2022 Track and Field World Championship

### 2023-04-18 Manuscript/manuscript.tex
Cleaned up some tables

%4
\jy{I think you have one objective but two methods to approach it.}
One of the first objectives of the paper is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
The model we use is a linear model, following a gamma distribution with a random
effect, which will be referred to as the year effect.  The year effect is
reflective of how significant the changes in the reaction time is for each venue,
i.e. the venue effect is  responsible for the variation between years.  
Another objective of this paper is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who
competed at the World Track and Field Championship in both 2019 and 2022 and
compare the times, and secondly we can look at how athletes performed compared
to how they did at their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

@@

\subsection{Beyond World Championship}
This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
\jy{cross-reference to the table by its label}
too quickly.  Table~\ref{USAvsWorld} below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. ``USA" denotes
the USA Track and Field Championships, ``World" denotes the World Track and Field
Championships, ``N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

@@

\subsection{GLMM}
%Include latex math symbols for the equations
The R package \citet{lme4} allows various linear mixed effects models to be
created, and was the primary modeling package used for our data analysis.
A gamma mixed linear effects model was chosen as the best model to model the data.
Many models were tried and compared to determine what the strongest model.  For
a fixed data set such as the men's pooled data including 2022, the models with
the venue effect included performed better than the models without the venue
effect for each set of comparable models; as indicated by the lower AIC and
higher log likelihood.  Thus, the venue effect is significant and should be
included in the model.  It can also be established that a gamma distribution
better models the data than linear data from comparing the linear model with
venue effect and them gamma model with a venue effect.  The final decision that
was made in model selection was to include link = log in the model.  Despite
doing very little to alter the AIC or log likelihood, the log scale was included
in order to try and better deal with high outliers. %Include Residuals plot?

Once the Gamma mixed effects model was chosen, it needed to be decided which combination
of effects to use.  It was possible to include a venue effect, a batch effect, 
or a year and batch effect.  The batch effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the batch reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the batch
effect, but are all factors that could potentially impact the venue effect.
Thus, we decided to inclue only the year effect in the GLMM, which is presented
as:

@@

\subsection{GLMM}
%Section 4.1: Gamma model results show the fitting with and without 2022,
%talk about p values and thresholds from the simulations (do not include the
%results in the methods section).

@@

At a size of $n=10,000,000$, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:
\jy{label the table and reference to it; clean it up}
\begin{center}
  \begin{tabular}{c c} 
   \toprule
   Data Set & Probability RxnTime < 0.1 \\ 
   \midrule
   all finals gamma log & 0.000856 \\ 
   old finals gamma log & 0.000439 \\
   all pooled gamma log & 0.006865 \\
   old pooled gamma log & 0.005761 \\
   \bottomrule
  \end{tabular}
  \end{center}

Using the gamma function with link set to log (whose summary statistics can be
found in the last row of the above table), the probability of the function being
less than 0.1 seconds can be calculated to see if Allen's disqualification was an anomaly.
Prior to the 2022 World Championship, there was a 0.0439\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals.  That means that we would expect a time below the threshold to occur
once every 2,277 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (The World Championship is biennial), every 569 years there
should be a time as extreme as Allen's.  This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

@@

\subsection{Rank-based Comparison}
%Section 4.2 is motivated by seiko timing and look at same athletes across
%multiple types of competition: 2019 and national

@@

As Seiko timed both events, there should not be a signficant
difference if the timing equipment in both instances was working correctly.
  To put it simply, there is not a reasonable explanation for 
this. It is inconceivable that so many athletes would improve such a significant
amount in such a small amount of time. 6 countries are represented in this 
comparison, which drastically decreases the chances of this issue being caused by
one specific country.  The United States athlete's improvement in times was
discussed earlier in the paper, but this data shows that on average there was
%Use a reference in this part
improvement for British, Polish, French, Brazilian, and Spanish athletes.

>>>>

%4
\jy{I think you have one objective but two methods to approach it.}
One of the first objectives of the paper is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
The model we use is a linear model, following a gamma distribution with a random
effect, which will be referred to as the venue effect.  The venue effect is
reflective of how significant the changes in the reaction time is for each venue,
i.e. the venue effect is  responsible for the variation between years.  
Another objective of this paper is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who
competed at the World Track and Field Championship in both 2019 and 2022 and
compare the times, and secondly we can look at how athletes performed compared
to how they did at their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

@@

\subsection{Beyond World Championship}
This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
\jy{cross-reference to the table by its label}
too quickly.  Table~\ref{fig:USAvsWorld} below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. ``USA" denotes
the USA Track and Field Championships, ``World" denotes the World Track and Field
Championships, ``N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

@@

\subsection{GLMM}
%Include latex math symbols for the equations
The R package lme3 allows various linear mixed effects models to be created, and 
was the primary modeling package used for our data analysis \citep{lme4}.
A gamma mixed linear effects model was chosen as the best model to model the data.
Many models were tried and compared to determine what the strongest model.  For
a fixed data set such as the men's pooled data including 2022, the models with
the venue effect included performed better than the models without the venue
effect for each set of comparable models; as indicated by the lower AIC and
higher log likelihood.  Thus, the venue effect is significant and should be
included in the model.  It can also be established that a gamma distribution
better models the data than linear data from comparing the linear model with
venue effect and them gamma model with a venue effect.  The final decision that
was made in model selection was to include link = log in the model.  Despite
doing very little to alter the AIC or log likelihood, the log scale was included
in order to try and better deal with high outliers. %Include Residuals plot?

Once the Gamma mixed effects model was chosen, it needed to be decided which combination
of effects to use.  It was possible to include a venue effect, a batch effect, 
or a year and batch effect.  The batch effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the batch reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the batch
effect, but are all factors that could potentially impact the venue effect.
Thus, we decided to inclue only the venue effect in the GLMM, which is presented
as:

@@

\subsection{GLMM} \label{subsec:Results_GLMM}
%Section 4.1: Gamma model results show the fitting with and without 2022,
%talk about p values and thresholds from the simulations (do not include the
%results in the methods section).

@@

As shown in table~\ref{fig:Gamma_parameters}, the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022.  The men's final data venue effect error increased from 0.439
to 0.504 and from 0.0495 to 0.0554 for the men's pooled data.  This demonstrates
that the inclusion of 2022 increased the randomness of the data and motivates
the rank-based study~\ref{subsec:Results_Rank}.

  Gamma model with venue effect and link=log
  \begin{center}
    \begin{tabular}{c  c  c  c  c  c  c } 
     \toprule
     Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
     \midrule
     Older Finals & 0.0439 & -1.901 & 0.0251 & 0.0019 & 0.0109 \\
     All Finals & 0.0504 & -1.914 & 0.0280 & 0.0025 & 0.0112 \\
     Older Pooled & 0.0495 & -1.853 & 0.0305 & 0.0024 & 0.0243 \\
     All Pooled & 0.0554 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\
     \bottomrule
     \label{fig:Gamma_parameters}
    \end{tabular}
    \end{center}

At a size of $n=10,000,000$, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:
\jy{label the table and reference to it; clean it up}
\begin{center}
  \begin{tabular}{c c} 
   \toprule
   Data Set & Probability RxnTime < 0.1 \\ 
   \midrule
   all finals gamma log & 0.000856 \\ 
   old finals gamma log & 0.000439 \\
   all pooled gamma log & 0.006865 \\
   old pooled gamma log & 0.005761 \\
   \bottomrule
   \label{fig:Sim_probability}
  \end{tabular}
  \end{center}

Using the gamma function with link set to log, the probability of the function being
less than 0.1 seconds can be calculated to see if Allen's disqualification was an anomaly.
Prior to the 2022 World Championship, there was a 0.0439\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals as shown in table~\ref{fig:Sim_probability}.
That means that we would expect a time below the threshold to occur
once every 2,277 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (The World Championship is biennial), every 569 years there
should be a time as extreme as Allen's.  This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

@@

\subsection{Rank-based Comparison} \label{subsec:Results_Rank}
%Section 4.2 is motivated by seiko timing and look at same athletes across
%multiple types of competition: 2019 and national

@@

As Seiko timed both events, there should not be a significant
difference if the timing equipment in both instances was working correctly.
  To put it simply, there is not a reasonable explanation for 
this. It is inconceivable that so many athletes would improve such a significant
amount in such a small amount of time. 6 countries are represented in this 
comparison, which drastically decreases the chances of this issue being caused by
one specific country.  The United States athlete's improvement in times was
discussed earlier in the paper, but this data shows that on average there was
%Use a reference in this part
improvement for British, Polish, French, Brazilian, and Spanish athletes.
